Thermodynamic response functions in a cell fluid model

### 2024-08-27 main_morse_response_functions.tex
remove const N where not needed

	\subsection{Pressure as a function of temperature and density}
	In this work, by density we mean the particle number density $\rho = \langle N \rangle / V$.
	In the framework of the grand canonical ensemble the average number of particles $\left\langle N \right\rangle$ is a partial derivative of natural logarithm of the grand partition function $\Xi$ with respect to the product $\beta \mu$. 
	\begin{eqnarray*}
		\langle N \rangle & = & \left(\frac{\partial \ln\Xi}{\partial \beta \mu}\right)_{T,V} 
		\\
		& = & \left(\frac{\partial (PV)}{\partial \mu}\right)_{T,V}
		\\
		& = & V \left(\frac{\partial P}{\partial \mu}\right)_{T,V}.
	\end{eqnarray*}
	From this equation it follows
	\begin{equation*}
		\frac{\langle N \rangle}{V} = \frac{\langle N \rangle}{v N_v} = \left(\frac{\partial P}{\partial \mu}\right)_{T,V}
	\end{equation*}
	and thus
	\begin{equation}
		\bar{n} \equiv \frac{\langle N \rangle}{V} v = \frac{\langle N \rangle}{N_v} = v \left(\frac{\partial P}{\partial \mu}\right)_{T,V} 
		= \left(\frac{\partial (Pv\beta)}{\partial (\beta \mu)}\right)_{T,V}.
	\end{equation}
	The quantity $\bar{n}$, on the one hand, has the meaning of the reduced particle number density. In the context of the cell model, on the other hand, it is the average number of particles per cell.
	Taking explicit derivatives, we arrive at
	\begin{equation}\label{eq:density}
	\bar n = n_g - M + \frac{ \bar \rho_0}{\beta W(0)}.
	\end{equation}
	The quantity $n_g$ in the equation \eqref{eq:density} is the critical density
	\begin{eqnarray}\label{eq:crit_dens}
	n_g & = & g_1 - \frac{g_3}{g_4}\left(  g_2 - \frac{1}{2} \frac{g_3^2}{g_4}\right) - \frac{1}{6} \frac{g_3^3}{g_4^2} 
	\nonumber\\
	& = & g_1 - \frac{g_2 g_3}{g_4} + \frac{g_3^3}{3g_4^2}
	\end{eqnarray}
	Its numerical value for parameters~\eqref{params} is
	\begin{equation*}
		n_g = 0.978.
	\end{equation*}
	Equations~\eqref{eq:density} and~\eqref{eq:eosMT} jointly define a parametric relationship between pressure and density, with $M$ serving as the parameter. Figure~\ref{fig1} shows the dependency of the reduced pressure $P^*$ on the density $\bar{n}$ for various values of the reduced temperature $\tau$, at and above the critical temperature.

@@

	\subsection{Isothermal compressibility}
	The isothermal compressibility is defined by
	\begin{equation}
		\label{def:isotherm_compres}
		K_T = -\frac{1}{V}\left(\frac{\partial V}{\partial P}\right)_{T,\langle N\rangle}
	\end{equation}
	Let us perform some transformations to rewrite the definition for $K_T$ into a form that is more suitable for the equation of state~\eqref{eq:eosMT}.
	\begin{eqnarray*}
		K_T & = & - \frac{1}{V}\left(\frac{\partial V}{\partial P}\right)_{T,\langle N\rangle}
		\\
		& = & -\rho \left(\frac{\partial \left(\frac{1}{\rho}\right)}{\partial P}\right)_{T,\langle N\rangle}
		\\
		& = & \frac{1}{\rho} \left(\frac{\partial \rho}{\partial P} \right)_{T,\langle N\rangle}
		\\
		& = & \frac{1}{\rho} \frac{\left(\partial \rho / \partial \mu\right)_{T,\langle N\rangle}}
		{\left(\partial P / \partial \mu\right)_{T,\langle N\rangle}},
	\end{eqnarray*}
	where $\rho = \langle N \rangle / V$ is the particle number density.
	Applying the Gibbs--Duhem equation
	\begin{equation}
		N{\rm d}\mu = -S{\rm d}T + V{\rm d}P,
	\end{equation}
	at $T = const$ one has 
	\begin{equation*}
		{\rm d} P = \rho {\rm d} \mu,
	\end{equation*}
	or
	\begin{equation}
		\label{eq:rho_dpdm}
		\rho = \left(\frac{\partial P}{\partial \mu}\right)_T.
	\end{equation}
	Substituting this into the last expression for $K_T$, one gets
	\begin{equation}
		K_T = \frac{1}{\rho^2} \left(\frac{\partial \rho}{\partial \mu}\right)_{T,\langle N\rangle}
	\end{equation}
	Finally, from~\eqref{eq:rho_dpdm} it follows that
	\begin{equation}
		\left(\frac{\partial \rho}{\partial \mu}\right)_T = \left(\frac{\partial^2 P}{\partial \mu^2}\right)_T
	\end{equation}
	and ultimately we arrive at the very useful expression for the isothermal compressibility
	\begin{equation}
		K_T = \frac{1}{\rho^2} \left(\frac{\partial^2 P}{\partial \mu^2}\right)_{T,\langle N\rangle}
	\end{equation}

	\begin{figure}[h!]
		\centering \includegraphics[width=0.5\textwidth]{f2a.pdf}
		\vskip-3mm\caption{The isothermal compressibility $K_T$ as a function of the density $\bar n$ at different values of reduced temperature $\tau > 0$ ($T > T_c$). 
		}\label{fig2a}
	\end{figure}

	\begin{figure}[h!]
		\centering \includegraphics[width=0.45\textwidth]{f2b.pdf}
		\includegraphics[width=0.475\textwidth]{f2c.pdf}
		\vskip-3mm\caption{The isothermal compressibility $K_T$ as a function of the effective chemical potential $M$ for different temperatures $\tau = (T - T_c)/T_c$ at $T > T_c$. The two figures differ in the scale of $M$. The left-hand-side figure covers a wider range of $M$, while the right-hand-side one focuses on a range of $M$ around its critical value $0$.
		}\label{fig2b}
	\end{figure}

>>>>

	\subsection{Pressure as a function of temperature and density}
	In this work, by density we mean the particle number density $\rho = \langle N \rangle / V$.
	In the framework of the grand canonical ensemble the average number of particles $\left\langle N \right\rangle$ is a partial derivative of natural logarithm of the grand partition function $\Xi$ with respect to $\beta \mu$ 
	\begin{eqnarray*}
		\langle N \rangle & = & \left(\frac{\partial \ln\Xi}{\partial \beta \mu}\right)_{T,V} 
		\\
		& = & \left(\frac{\partial (PV)}{\partial \mu}\right)_{T,V}
		\\
		& = & V \left(\frac{\partial P}{\partial \mu}\right)_{T,V}.
	\end{eqnarray*}
	From this equation it follows
	\begin{equation*}
		\frac{\langle N \rangle}{V} = \frac{\langle N \rangle}{v N_v} = \left(\frac{\partial P}{\partial \mu}\right)_{T,V}
	\end{equation*}
	and thus
	\begin{equation}
		\bar{n} \equiv \frac{\langle N \rangle}{V} v = \frac{\langle N \rangle}{N_v} = v \left(\frac{\partial P}{\partial \mu}\right)_{T,V} 
		= \left(\frac{\partial (Pv\beta)}{\partial (\beta \mu)}\right)_{T,V}.
	\end{equation}
	The quantity $\bar{n}$, on the one hand, has the meaning of the reduced particle number density. In the context of the cell model, on the other hand, it is the average number of particles per cell.
	Taking explicit derivatives, we arrive at
	\begin{equation}\label{eq:density}
	\bar n = n_g - M + \frac{ \bar \rho_0}{\beta W(0)}.
	\end{equation}
	The quantity $n_g$ in the equation \eqref{eq:density} is the critical density
	\begin{eqnarray}\label{eq:crit_dens}
	n_g & = & g_1 - \frac{g_3}{g_4}\left(  g_2 - \frac{1}{2} \frac{g_3^2}{g_4}\right) - \frac{1}{6} \frac{g_3^3}{g_4^2} 
	\nonumber\\
	& = & g_1 - \frac{g_2 g_3}{g_4} + \frac{g_3^3}{3g_4^2}
	\end{eqnarray}
	Its numerical value for parameters~\eqref{params} is
	\begin{equation*}
		n_g = 0.978.
	\end{equation*}
	Equations~\eqref{eq:density} and~\eqref{eq:eosMT} jointly define a parametric relationship between pressure and density, with $M$ serving as the parameter. Figure~\ref{fig1} shows the dependency of the reduced pressure $P^*$ on the density $\bar{n}$ for various values of the reduced temperature $\tau$, at and above the critical temperature.

@@

	\subsection{Isothermal compressibility}
	The isothermal compressibility is defined by
	\begin{equation}
		\label{def:isotherm_compres}
		K_T = -\frac{1}{V}\left(\frac{\partial V}{\partial P}\right)_{T,\langle N\rangle}.
	\end{equation}
	Let us perform some transformations to rewrite $K_T$ into a form that is more suitable for the equation of state~\eqref{eq:eosMT}
	\begin{eqnarray*}
		K_T & = & - \frac{1}{V}\left(\frac{\partial V}{\partial P}\right)_{T,\langle N\rangle}
		\\
		& = & -\rho \left(\frac{\partial \left(\frac{1}{\rho}\right)}{\partial P}\right)_{T,\langle N\rangle}
		\\
		& = & \frac{1}{\rho} \left(\frac{\partial \rho}{\partial P} \right)_{T,\langle N\rangle}
		\\
		& = & \frac{1}{\rho} \frac{\left(\partial \rho / \partial \mu\right)_{T}}
		{\left(\partial P / \partial \mu\right)_{T}},
	\end{eqnarray*}
	where $\rho = \langle N \rangle / V$ is the particle number density. We have omitted the condition of constant $\langle N \rangle$ in the last line of the above equation since we have explicit dependencies on temperature and chemical potential for both pressure $P = P(T, \mu)$, see Eq.~\eqref{eq:eosMT}, and density $\rho = \rho(T, \mu)$, see Eq.~\eqref{eq:density}.
	Applying the Gibbs--Duhem equation
	\begin{equation}
		N{\rm d}\mu = -S{\rm d}T + V{\rm d}P,
	\end{equation}
	at $T = const$ one has 
	\begin{equation*}
		{\rm d} P = \rho {\rm d} \mu,
	\end{equation*}
	or
	\begin{equation}
		\label{eq:rho_dpdm}
		\rho = \left(\frac{\partial P}{\partial \mu}\right)_T.
	\end{equation}
	Substituting this into the last expression for $K_T$, one gets
	\begin{equation}
		K_T = \frac{1}{\rho^2} \left(\frac{\partial \rho}{\partial \mu}\right)_{T}
	\end{equation}
	Finally, from~\eqref{eq:rho_dpdm} it follows that
	\begin{equation}
		\left(\frac{\partial \rho}{\partial \mu}\right)_T = \left(\frac{\partial^2 P}{\partial \mu^2}\right)_T
	\end{equation}
	and ultimately we arrive at the very useful expression for the isothermal compressibility
	\begin{equation}
		K_T = \frac{1}{\rho^2} \left(\frac{\partial^2 P}{\partial \mu^2}\right)_{T}
	\end{equation}

	Let us introduce the reduced isothermal compressibility
	\begin{equation}
		K^*_T \equiv \frac{\varepsilon K_T}{v}.
	\end{equation}
	The quantity $K^*_T$ is dimensionless and is of order unity, except at the critical point itself, where it is divergent. Figure~\ref{fig2a} illustrates the dependency of the compressibility $K^*_T$ on the density $\bar{n}$ for various values of temperature above the critical one.

	\begin{figure}[h!]
		\centering \includegraphics[width=0.5\textwidth]{f2a.pdf}
		\vskip-3mm\caption{The reduced isothermal compressibility $K^*_T$ as a function of the density $\bar n$ at different values of reduced temperature $\tau > 0$ ($T > T_c$). 
		}\label{fig2a}
	\end{figure}

	\begin{figure}[h!]
		\centering \includegraphics[width=0.45\textwidth]{f2b.pdf}
		\includegraphics[width=0.475\textwidth]{f2c.pdf}
		\vskip-3mm\caption{The reduced isothermal compressibility $K^*_T$ as a function of the effective chemical potential $M$ for different temperatures $\tau = (T - T_c)/T_c$ at $T > T_c$. The two figures differ in the scale of $M$. The left-hand-side figure covers a wider range of $M$, while the right-hand-side one focuses on a range of $M$ around its critical value $0$.
		}\label{fig2b}
	\end{figure}
